Racial and Ethnic Differences in ETI

### 2023-09-15 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
back to gtsummary for race and ETI

---
title: "Racial and Ethnic Differences in ETI"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /Users/timvigers/Documents/Miscellaneous/zotero.bib
csl: /Users/timvigers/Documents/Miscellaneous/american-medical-association.csl
editor: source
---

```{r libraries}
#| include: false
library(tidyverse)
library(table1)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

@@

```{r}
table1(~race, data = demo)
```

@@

```{r}
table1(~ race | factor(ReviewYear), data = annual)
```

@@

```{r}
t1_vars <- c(
  "Gender", "WasGenotyped", "F508_num", "mean_ppFEV", "Age_YrEnd",
  "mean_ppFEV", "weight_last", "weight_perc_last", "bmi_last",
  "bmi_perc_last", "Vx770", "Vx809comb", "Vx661comb", "Vx445comb"
)
t1_form <- as.formula(paste("~", paste(t1_vars, collapse = "+"), "|factor(ReviewYear)"))
```

@@

```{r}
table1(~ Gender + WasGenotyped + F508_num | race, data = demo)
```

@@

```{r}
table1(t1_form, data = annual)
```

@@

```{r}
table1(~ eti_elig | factor(ReviewYear), data = annual, overall = F)
```

@@

```{r}

```

@@

```{r}
encounter %>%
  filter(encounterdate >= "2019-01-01") %>%
  mutate(post_covid = encounterdate > "2020-03-11") %>%
  group_by(eDWID) %>%
  summarise(
    `Number of Visits Pre-COVID` = sum(!post_covid),
    `Number of Visits Post-COVID` = sum(post_covid),
    race = first(race),
    .groups = "drop"
  ) %>%
  rename(any_of(tidy_labels)) %>%
  select(-eDWID) %>%
  tbl_summary(by = `Race/Ethnicity`, missing_text = "Missing") %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
mod_df <- encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(
    across(
      c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted),
      ~ mean(.x, na.rm = T)
    ),
    race = first(race),
    .groups = "drop"
  )
```

@@

```{r}
#| warning: false
mod_df
pivot_longer(c("GLI_FEV1_pct_predicted", "GLI_FVC_pct_predicted")) %>%
  ggplot(aes(x = reviewyear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  GLI_FEV1_pct_predicted ~ factor(reviewyear) * race + (1 | eDWID),
  data = mod_df
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + reviewyear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```

```{r}
#| fig-width: 10
#| fig-height: 10
ggplot(means, aes(x = reviewyear, y = emmean, color = race)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = .2) +
  theme_bw()
```

>>>>

---
title: "Racial and Ethnic Differences in ETI"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /Users/timvigers/Dropbox/Miscellaneous/zotero.bib
csl: /Users/timvigers/Dropbox/Miscellaneous/american-medical-association.csl
editor: source
---

```{r setup}
#| include: false
# Libraries
library(Hmisc)
library(tidyverse)
library(gtsummary)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

@@

```{r}
demo %>%
  select(race) %>%
  tbl_summary(missing_text = "Missing") %>%
  bold_labels()
```

@@

```{r}
annual %>%
  select(race, ReviewYear) %>%
  tbl_summary(by = ReviewYear, missing_text = "Missing") %>%
  bold_labels()
```

@@

```{r}
t1_vars <- c(
  "Gender", "WasGenotyped", "F508_num", "mean_ppFEV", "Age_YrEnd",
  "mean_ppFEV", "weight_last", "weight_perc_last", "bmi_last",
  "bmi_perc_last", "Vx770", "Vx809comb", "Vx661comb", "Vx445comb"
)
```

@@

```{r}
demo %>%
  select(Gender, WasGenotyped, F508_num, race) %>%
  tbl_summary(by = race, missing_text = "Missing") %>%
  bold_labels()
```

@@

```{r results='asis'}
set_gtsummary_theme(theme_gtsummary_compact(), quiet = T)
years <- split.data.frame(annual, annual$ReviewYear)
invisible(lapply(names(years), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(all_of(t1_vars), race) %>%
    tbl_summary(by = race, missing_text = "Missing") %>%
    bold_labels()
  print(t)
  cat("\n")
}))
reset_gtsummary_theme()
```

@@

```{r}
annual %>%
  select(eti_elig, ReviewYear) %>%
  tbl_summary(
    by = ReviewYear, missing_text = "Missing",
    type = all_dichotomous() ~ "categorical"
  ) %>%
  bold_labels()
```

@@

```{r results='asis'}
invisible(lapply(names(years), function(n) {
  cat("\n")
  cat("###", n)
  cat("\n")
  d <- years[[n]]
  t <- d %>%
    select(eti_elig, race) %>%
    tbl_summary(
      by = race, missing_text = "Missing",
      type = all_dichotomous() ~ "categorical"
    ) %>%
    bold_labels()
  print(t)
  cat("\n")
}))
```

# People without visits in a given year

```{r}
# Which are missing from encounters data?
a <- as.data.frame.matrix(table(annual$eDWID, annual$ReviewYear))
a <- a %>% rownames_to_column("eDWID")
e <- as.data.frame.matrix(table(by_year$eDWID, by_year$reviewyear))
e <- e %>% rownames_to_column("eDWID")
rowdiff_a <- setdiff(a, e)
rowdiff_a$data <- "annual"
rowdiff_e <- setdiff(e, a)
rowdiff_e$data <- "encounter"
diffs <- rbind(rowdiff_a, rowdiff_e)
diffs <- diffs %>%
  arrange(eDWID, desc(data)) %>%
  select(-data)
diffs <- diffs %>%
  group_by(eDWID) %>%
  reframe(across(`2015`:`2021`, ~ diff(as.numeric(.x))))
diffs %>%
  reframe(across(`2015`:`2021`, ~ sum(.x == 1))) %>%
  kable()
```

There were a total of `r length(setdiff(a$eDWID,e$eDWID))` included in the annual dataset that did not appear in the encounter-level dataset at all. There were an additional `r nrow(diffs)` patients who were included in both datasets, but had at least one year with no encounters. The table above shows the number of people missing an encounter in each year.

If someone was not seen a given year, do we count them as eligible for that year? I am assuming they will be eligible but obviously will not have started any modulators.

@@

```{r}
encounter %>%
  filter(encounterdate >= "2019-01-01") %>%
  mutate(post_covid = encounterdate > "2020-03-11") %>%
  group_by(eDWID) %>%
  summarise(
    `Number of Visits Pre-COVID` = sum(!post_covid),
    `Number of Visits Post-COVID` = sum(post_covid),
    race = first(race),
    .groups = "drop"
  ) %>%
  select(-eDWID) %>%
  tbl_summary(by = race, missing_text = "Missing") %>%
  bold_labels()
```

@@

@@

```{r}
#| warning: false
annual %>%
  mutate(mean_ppFEV = as.numeric(mean_ppFEV), mean_ppFVC = as.numeric(mean_ppFVC)) %>%
  pivot_longer(c("mean_ppFEV", "mean_ppFVC")) %>%
  ggplot(aes(x = ReviewYear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  mean_ppFEV ~ factor(ReviewYear) * race + (1 | eDWID),
  data = annual
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + ReviewYear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```

```{r}
ggplot(means, aes(x = ReviewYear, y = emmean, color = race)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("LS Mean FEV1") +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = .2) +
  theme_bw()
```
